---
output:
    github_document:
      html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mikropml

> meek-ROPE em el

<!-- badges: start -->
[![check](https://github.com/SchlossLab/mikropml/workflows/check/badge.svg)](https://github.com/SchlossLab/mikropml/actions?query=workflow%3Acheck+branch%3Amaster)
[![codecov](https://codecov.io/gh/SchlossLab/mikropml/branch/master/graph/badge.svg)](https://codecov.io/gh/SchlossLab/mikropml)
[![docs](https://img.shields.io/static/v1?style=flat&label=Docs&message=Website&color=success)](http://www.schlosslab.org/mikropml/)
[![license](https://img.shields.io/badge/license-MIT-blue.svg)](https://github.com/SchlossLab/mikropml/blob/master/LICENSE.md)
[![lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![CRAN](https://img.shields.io/cran/v/mikropml?color=blue&label=CRAN&logo=R)](https://CRAN.R-project.org/package=mikropml)
<!-- badges: end -->

User-Friendly R Package for Supervised Machine Learning Pipelines

## Installation

You can install the latest release from CRAN:
```{r, eval = FALSE}
install.packages('mikropml')
```

or the development version from 
[GitHub](https://github.com/SchlossLab/mikRopML):

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SchlossLab/mikropml")
```

### Dependencies

```{r, echo = FALSE, message = FALSE}
library(dplyr)
description <- utils::packageDescription('mikropml', 
                                         fields = c('Imports', 'Suggests'))
deps <- lapply(names(description), 
               function (x) {
                 paste0('- ', x, ': ', 
                        description[[x]] %>% 
                          gsub("\n", " ", .))}
               ) %>% 
  unlist() %>% 
  paste(., collapse = '\n')
```

`r deps`

## Usage

Check out the 
[introductory vignette](http://www.schlosslab.org/mikropml/articles/introduction.html) 
for a quick start tutorial. 
For a more in-depth discussion, read 
[all the vignettes](http://www.schlosslab.org/mikropml/articles/index.html) 
and/or take a look at the 
[reference documentation](http://www.schlosslab.org/mikropml/reference/index.html).

## Help & Contributing

If you come across a bug, [open an issue](https://github.com/SchlossLab/mikropml/issues) and include a [minimal reproducible example](https://www.tidyverse.org/help/).

If you'd like to contribute, see our guidelines [here](http://www.schlosslab.org/mikropml/CONTRIBUTING.html).

## Code of Conduct

Please note that the mikropml project is released with a [Contributor Code of Conduct](http://www.schlosslab.org/mikropml/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Why the name?

The word "mikrop" (pronounced "meek-ROPE") is Turkish for "microbe".
This package was originally implemented as a machine learning pipeline for microbiome-based classification problems (see [Topçuoğlu _et al._ 2020](https://doi.org/10.1128/mBio.00434-20)).
We realized that these methods are applicable in many other fields too,
but stuck with the name because we like it!
