---
title: "Parallel processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r setup}
library(mikropml)
library(tictoc)
```

## Speeding up `run_ml()`

By default, the `run_ml()` function only uses one core.
```{r series}
tic()
result1 <- run_ml(otu_small, 'regLogistic', seed = 2019)
toc()
```

If you'd like to parallelize the cross-validation step with multiple cores,
use the `setup_parallel()` function to specify how many cores you have available
prior to calling `run_ml()`.
```{r parallel}
setup_parallel(2)
tic()
result2 <- run_ml(otu_small, 'regLogistic', seed = 2019)
toc()
```

`setup_parallel()` uses the `future::multiprocess` plan to split the work across multiple cores.
See the [`doFuture` documentation](https://cran.r-project.org/web/packages/doFuture/vignettes/doFuture.html) for more details or if you'd like to use a different futures plan.
You will need to have the `foreach`, `future`, `doFuture`, and `parallel` packages installed to use this feature.
Optionally, you can also install the `future.apply` package to speed up other
various steps in `preprocess_data()` and `run_ml()`.

## Running multiple calls to `run_ml()` in parallel
